Writing Trash Bin

### 2024-04-01 paper.qmd
removed old paragraphs after Di edit

<!-- TODO 
- convert to American spelling
- should it be data-driven or two words
- May not need to reference the same paper again and again
-->

@@

<!-- ## Introduction-->
From entertainment choices to news articles to insurance plans, the modern citizen is so over run with information in every aspect of their life it can be overwhelming. In this overflow of information, tools that can effectively summarise information down into simple and clear ideas become more valuable. Information visualisations remain one of the most powerful tools for fast and reliable science communication.

There are many stages in our analysis that might benefit from the power of data visualisation, however this does not mean it is appropriately utilized or done with with effectiveness. (@Vanderplas2020 has helpful advice here.) Visualization is an important step in exploratory data analysis and it is often utilised to **learn** what is important about a data set. The importance of data-driven discovery has been highlighted through the data sets Anscombe's quartet [@anscombe] and the Datasaurus Dozen [@datasaurpkg]. Each of the pairwise plots in these data sets have the same summary statistics but strikingly different information, as revealed when visualised. <!-- Anscombe quartet is shown in @fig-anscombe, because describing the data is never the same as seeing it. Instead of having to repeatedly check endless hypothesis to find interesting numerical features: DON'T NEED MUCH EMPHASIS ON THIS DATA - VERY WELL-KNOWN -->
Visualisations can **tell** us what is important about our data and **how** it might diverge from pre-conceived notions or what we expect.

```{r}
#| echo: false
#| warning: false
#| message: false
#| eval: false
#| label: fig-anscombe
#| fig-cap: "The four scatter plots that make up Anscombe's quartet. The four scatter plots are visually distinct but have the same mean, standard deviation, and correlation. The visualisation highlights the importance of plotting your data to identify interesting features that are hidden by other summary statistics."
library(tidyverse)
tibble(x = c(anscombe$x1, anscombe$x2,
             anscombe$x3, anscombe$x4),
       y = c(anscombe$y1, anscombe$y2,
             anscombe$y3, anscombe$y4),
       Plot = c(rep("Plot 1",11), rep("Plot 2",11), 
                rep("Plot 3",11), rep("Plot 4",11))) %>%
  ggplot(aes(x,y)) +
  geom_point(aes(fill=Plot), colour="black", 
             size=3, pch=21, alpha=0.75) +
  facet_wrap(~Plot) +
  theme_classic() +
  theme(aspect.ratio = 1,
        legend.position = "none") +
  scale_fill_brewer(type = "qual", palette = 4)
```

Uncertainty visualisation is a relatively new field in research. Early papers that specifically reference "uncertainty visualisation" appear in the late 80s [@Ibrekk1987], with geospatial information visualisation literature in the early 90s declaring this to be essential aspect of information display [@MacEachren1992; @Carr1992]. Despite the new terminology visualisation of uncertainty has been present since the earliest times. For example, box plots or histograms can be considered to be displaying uncertainty in the sense of variability in observations sampled from a population distribution. <!--In the modern day, uncertainty visualisation has exploded into its own field that applies to a wide range of contexts.-->

<!-- ## Similar work-->

<!-- Any field that experiences an explosion of new research will also invite a series of literature reviews that seek to combine and summaries that information. Given that uncertainty visualisation has been around for a few decades now, there is already a wealth of uncertainty visualisation literature reviews. -->

@@

Interestingly, these reviews often do not offer overarching rules for tried and tested uncertainty visualisation, but rather comment on the *difficulties* faced when trying to summarise the papers from this field. @Kinkeldey2014 found most experiments on the methods for uncertainty visualisation evaluation to be ad hoc, with no commonly agreed upon methodology or formalisation and no greater goal of describing general principals. @Hullman2016 commented on the difficulty in taking overarching themes from uncertainty visualisation, as several conflated issues make it unclear if subjects did poorly in an experiment because they misunderstood a visualisation, because the question was misinterpreted, or because they used a specific heuristic. @Spiegelhalter2017 commented that different plots are good for different things, and disagreed with the goal of identifying a universal "best" plot for all people and circumstances. @Griethe2006 was unable to find common themes, but instead listed the findings and opinions of <!--each uncertainty visualisation--> a collection of papers.

There are several reasons provided to explain the difficulty in generalising <!--the findings in--> the uncertainty visualisation literature. Some suggested or agree with the idea that visualisation typologies should move away from data types, uncertainty categories, and representation types and towards "task-centred typologies" [@Kinkeldey2014; @Hullman2016]. Other papers indirectly hint at this <!--Some authors may not attribute the task to the *reason* the literature cannot be easily summarised, but do mention that--> by arguing that the choice of best visualisation is highly dependent on the specific goal [@Griethe2006; @Spiegelhalter2017]. Others contend that visualisations are highly dependent on the audience and there is no such thing as a "best" visualisation that will be accessible to all members [@Kinkeldey2014; @Spiegelhalter2017]. These concerns appear <!--difficulties in noisy uncertainty visualisation findings arises regardless of the direction the analysis was approached, be it looking--> across all application areas,  <!--, in geospatial data [@Kinkeldey2014], or uncertainty visualisation as a whole [@Hullman2016; @Spiegelhalter2017; @Griethe2006].--> and an underlying thread is that the problem arises from a lack of a cohesive and encompassing **definition of uncertainty**.

<!--This review believe all these sources of noise have a single root cause, a lack of a cohesive and encompassing definition of uncertainty. The absence of an encompassing definition of uncertainty is mentioned by every uncertainty visualisation directly [@Spiegelhalter2017; @Griethe2006] or indirectly by describing a myriad of ways it can be considered in the literature [@Kinkeldey2014; @Hullman2016], although it is never commented on as a source of the noise in the field.-->

The review provided here attempts to address these issues. We begin with understanding the definition problem, to obtain and understanding of the *purpose* of visualising uncertainty, and understand if the current visualisation tools are sufficient to achieve that purpose and if not, where there is need for new methodology. We also posit that "uncertainty visualisation" is too diffuse to be considered a field in itself, but rather should be dissected into smaller elements. The ideal role of this review to help synthesize this noisy field, by collecting together essential components, summarising different viewpoints the current literature, and provide a guide for potential research topics. <!-- In understanding this purpose, we will also consider whether or not "uncertainty visualisation" is a valid sub-field of visualisation itself, or if that distinction has only arisen as a by product of the confused definition of uncertainty. Ultimately, this paper should serve as a guide for future uncertainty visualisation authors to prevent their work from becoming noise in what is already an incredibly noisy field.-->

# What is Uncertainty?
<!--## Colloquial definitions - what's in this section is not colloquial definitions, it's more overview-->

Uncertainty visualisation is directed towards communication - to the general public when disasters may hit, to decision makers in control centers, and a whole host of other public-facing tasks - needing an understanding of what it is when the term "uncertainty" is used. <!--is made particularly difficult by the term "uncertainty" lacking a commonly accepted definition in the literature. This mishmash of terminology leads to a large body of work, all claiming to finding the best visualisation or expression of of "uncertainty" but most don't even seem to agree on what uncertainty is. 
-->
Definitions in the literature appear seemingly view it to be encompassing and include everything the general public might consider to be uncertain. Some works (XXX eg) focus more narrowly on specific terms defined mathematically, such as probability, variance, error, or precision. Others (XXX eg)  <!--consider it to be an encompassing umbrella term of which for these mathematically defined objects are only examples, and-->  include broader loosely related elements, such as missing values (*Cite Gap 3: Organise paper definitions*). <!--This wide array of uncertainty definitions becomes a snake eating its own tail, as many authors opt to redefine uncertainty themselves[], ignore defining uncertainty entirely [], or pick a definition from a seemingly randomly previously published papers [], all of which make variance in uncertainty definitions even worse (*Cite Gap 3: Organise paper definitions*).

While these vague definitions might be OK for polite conversation, they are *not* foundation on which to build an entire sub-field of visualisation. In order to visualise uncertainty, it needs to be quantifiable [@Griethe2006; Leland2005], in order to quantify uncertainty, it needs to be mathematically defined. The uncertainty visualisation literature is completely awash with papers that define uncertainty using a vague encompassing phrase, (e.g. @utypo defined it as any deviation from complete determinism) but go on to *quantify* uncertainty as a PDF, error, or some other easily quantified mathematical object when it comes time to do the visualisation (*Cite Gap 4: Organise uncertainty quantification*). This definition swap happens so subtly that it seems to go unnoticed by authors writing the papers, and the fact that "uncertainty visualisation" methods don't even notice they are exclusively focus on depictions of easily quantifiable uncertainty, such as error [], mass [], or variance [] (*Cite Gap 4: Organise uncertainty quantification*). This has left the expressions of uncertainty that are hard to quantify or visually differentiate untouched, despite many papers calling for their invention (*Cite Gap 6*). There is likely only confusion and obfuscation if the vague definitions of "uncertainty" continue to be used and authors should try to be more specific with the mathematical object they are visualising, *or* a strict definition of uncertainty needs to be defined. The current method of titling papers and defining uncertainty forces readers to comb through the method of a particular visualisation experiment to understand if the uncertainty visualised is amenable to their needs. 
-->
Several attempts to address this gap are made in the papers describing taxonomies. 
<!-- Some papers have recognised these colloquial definitions are lacking and have made attempts to formalise the space with taxonomies, however these taxonomies come with their own issues. -->

## Definitions in taxonomies
<!--Much in the same way that there are almost as many definitions of uncertainty as there are papers on the topic, the field is also over run with taxonomies.--> Taxonomies split uncertainty based on an endless stream of ever changing boundaries, such as whether the uncertainty is due to true randomness or a lack of knowledge [@Spiegelhalter2017], if the uncertainty is in the  attribute, spatial elements, or temporal element of the data [], whether the uncertainty is scientific (e.g. error) or human (e.g. disagreement among parties) [], if the uncertainty is random or systematic [], statistical or bounded [], accuracy or precision [], if the uncertainty is about the past or the future [], the stage of the data analysis pipeline it comes from [], the severity of the uncertainty [], how quantifiable the uncertainty is [@Spiegelhalter2017], etc., etc., etc. (*Cite Gap 7: Uncertainty Taxonomies*). In their own way, each of these taxonomies show an aspect of uncertainty that an author felt was important to differentiate. @utypo identified many common themes among this wide array of taxonomies and used them to design an encapsulating taxonomy, depicted in @fig-taxonomy. This definition organises uncertainty based on three axis, location (where the uncertainty is coming from), level (how quantifiable it is), and nature (if it comes from true randomness or incomplete knowledge).

@@

# Future work
This paper has identified gaps in the uncertainty visualisation literature that could be filled to progress the field. <!--and with those issues, we have several suggestions for future work that would likely be fruitful for the field.-->

<!--Our first suggestion for future work is--> *Each new development should be accompanied by a mathematical definition of the uncertainty being addressed.* Utilizing bias and variance to define uncertainty is well-established, but it is absent concerns about imputed data, assumptions, sampling methods, and even the human element of analyst decisions. <!--There is currently a lot of work that quantifies the uncertainty through bias and variance at various stages of the data analysis pipeline separately. Elements such as  imputed data, assumptions, sampling methods, and analyst decisions all have their own independent work, quantification and discussions, but methods to brand between these concepts are few and far between.--> A conceptual framework that allows us to combine these components is needed.

*The concept of uncertainty should be formalised within the grammar of graphics.* This formalisation would allow uncertainty visualisation authors to have a clear understanding of what is or is not an uncertainty visualisation. Additionally placing uncertainty visualisation in the framework that is used to understand existing information visualisation research would help authors understand when existing methods can be used to explain their results. incorporating uncertainty into the grammar of graphics will also give a more precise concept of the information contained within a plot. Other fields of science employ marginal changes when designing experiments to ensure it is well understood *what* aspect of their experiment is contributing to their results, and a better sense of what "marginal" is in the case of uncertainty visualisation would greatly help the field.

XXX Is data pipeline connected with the grammar of graphics? Should this be a recommendation?

@@

If an uncertainty visualisation researcher would prefer to perform experiments rather than formalise methods, there are options there too. It would be interesting to know if any perceptual tasks that can be mapped to two different visual tasks condense into a single dimension when looking for overarching signal in a plot. Alternatively, the task dependency many authors in uncertainty visualisation mention would be a useful direction to consider. It is clear that the the number of potential tasks that can be performed on a visualisation increases with with the number of observations. A single observation is limited to value extraction, two observations can be compared, multiple observations allow for shapes or global statistics to be extracted. The interaction between sample size and task is of particular interest to the uncertainty visualisation community, as uncertainty can be expressed through multiple observations using a sample, or through a single value using an error. Of course, this is limited by the fact that there also isn't a definition for what is a "task" and given the mess created by the lack of formalisation in uncertainty visualisation, it may be wise to formalise that concept before performing these experiments. @Amar2005 suggested a taxonomy for information visualisation based on the types of tasks we use visualisations for and suggest 10 "analytical primitives" that we can then map to visualisations, which could be a good starting point. Regardless, these are directions of research would be fruitful to the uncertainty visualisation community even if it appears on the surface to be research that is only beneficial to the "normal" visualisation community. XXX Not sure what this paragraph is recommending?

>>>>

<!-- TODO 
- convert to American spelling
- May not need to reference the same paper again and again
-->

@@

From entertainment choices to news articles to insurance plans, the modern citizen is so over run with information in every aspect of their life it can be overwhelming. In this overflow of information, tools that can effectively summarise information down into simple and clear ideas become more valuable. Information visualisations remain one of the most powerful tools for fast and reliable science communication.

There are many stages in our analysis that might benefit from the power of data visualisation, however this does not mean it is appropriately utilized or done with with effectiveness. (@Vanderplas2020 has helpful advice here.) Visualization is an important step in exploratory data analysis and it is often utilised to **learn** what is important about a data set. The importance of data-driven discovery has been highlighted through the data sets Anscombe's quartet [@anscombe] and the Datasaurus Dozen [@datasaurpkg]. Each of the pairwise plots in these data sets have the same summary statistics but strikingly different information, as revealed when visualised. Visualisations can **tell** us what is important about our data and **how** it might diverge from pre-conceived notions or what we expect.

Uncertainty visualisation is a relatively new field in research. Early papers that specifically reference "uncertainty visualisation" appear in the late 80s [@Ibrekk1987], with geospatial information visualisation literature in the early 90s declaring this to be essential aspect of information display [@MacEachren1992; @Carr1992]. Despite the new terminology visualisation of uncertainty has been present since the earliest times. For example, box plots or histograms can be considered to be displaying uncertainty in the sense of variability in observations sampled from a population distribution.

@@

Interestingly, these reviews often do not offer overarching rules for tried and tested uncertainty visualisation, but rather comment on the *difficulties* faced when trying to summarise the papers from this field. @Kinkeldey2014 found most experiments on the methods for uncertainty visualisation evaluation to be ad hoc, with no commonly agreed upon methodology or formalisation and no greater goal of describing general principals. @Hullman2016 commented on the difficulty in taking overarching themes from uncertainty visualisation, as several conflated issues make it unclear if subjects did poorly in an experiment because they misunderstood a visualisation, because the question was misinterpreted, or because they used a specific heuristic. @Spiegelhalter2017 commented that different plots are good for different things, and disagreed with the goal of identifying a universal "best" plot for all people and circumstances. @Griethe2006 was unable to find common themes, but instead listed the findings and opinions of a collection of papers.

There are several reasons provided to explain the difficulty in generalising the uncertainty visualisation literature. Some suggested or agree with the idea that visualisation typologies should move away from data types, uncertainty categories, and representation types and towards "task-centred typologies" [@Kinkeldey2014; @Hullman2016]. Other papers indirectly hint at this by arguing that the choice of best visualisation is highly dependent on the specific goal [@Griethe2006; @Spiegelhalter2017]. Others contend that visualisations are highly dependent on the audience and there is no such thing as a "best" visualisation that will be accessible to all members [@Kinkeldey2014; @Spiegelhalter2017]. These concerns appear across all application areas, and an underlying thread is that the problem arises from a lack of a cohesive and encompassing **definition of uncertainty**.

The review provided here attempts to address these issues. We begin with understanding the definition problem, to obtain and understanding of the *purpose* of visualising uncertainty, and understand if the current visualisation tools are sufficient to achieve that purpose and if not, where there is need for new methodology. We also posit that "uncertainty visualisation" is too diffuse to be considered a field in itself, but rather should be dissected into smaller elements. The ideal role of this review to help synthesize this noisy field, by collecting together essential components, summarising different viewpoints the current literature, and provide a guide for potential research topics.

# What is Uncertainty?
Uncertainty visualisation is directed towards communication - to the general public when disasters may hit, to decision makers in control centers, and a whole host of other public-facing tasks - needing an understanding of what it is when the term "uncertainty" is used.

Definitions in the literature appear seemingly view it to be encompassing and include everything the general public might consider to be uncertain. Some works (*XXX eg*) focus more narrowly on specific terms defined mathematically, such as probability, variance, error, or precision. Others (*XXX eg*) include broader loosely related elements, such as missing values. Several attempts to address this gap are made in the papers describing taxonomies.

## Definitions in taxonomies
Taxonomies split uncertainty based on an endless stream of ever changing boundaries, such as whether the uncertainty is due to true randomness or a lack of knowledge [@Spiegelhalter2017], if the uncertainty is in the  attribute, spatial elements, or temporal element of the data [], whether the uncertainty is scientific (e.g. error) or human (e.g. disagreement among parties) [], if the uncertainty is random or systematic [], statistical or bounded [], accuracy or precision [], if the uncertainty is about the past or the future [], the stage of the data analysis pipeline it comes from [], the severity of the uncertainty [], how quantifiable the uncertainty is [@Spiegelhalter2017], etc., etc., etc. (*Cite Gap 7: Uncertainty Taxonomies*). In their own way, each of these taxonomies show an aspect of uncertainty that an author felt was important to differentiate. @utypo identified many common themes among this wide array of taxonomies and used them to design an encapsulating taxonomy, depicted in @fig-taxonomy. This definition organises uncertainty based on three axis, location (where the uncertainty is coming from), level (how quantifiable it is), and nature (if it comes from true randomness or incomplete knowledge).

@@

# Future work
This paper has identified gaps in the uncertainty visualisation literature that could be filled to progress the field.

*Each new development should be accompanied by a mathematical definition of the uncertainty being addressed.* Utilizing bias and variance to define uncertainty is well-established, but it is absent concerns about imputed data, assumptions, sampling methods, and even the human element of analyst decisions. A conceptual framework that allows us to combine these components is needed.

*The concept of uncertainty should be formalised within the grammar of graphics.* This formalisation would allow uncertainty visualisation authors to have a clear understanding of what is or is not an uncertainty visualisation. Additionally placing uncertainty visualisation in the framework that is used to understand existing information visualisation research would help authors understand when existing methods can be used to explain their results. incorporating uncertainty into the grammar of graphics will also give a more precise concept of the information contained within a plot. Other fields of science employ marginal changes when designing experiments to ensure it is well understood *what* aspect of their experiment is contributing to their results, and a better sense of what "marginal" is in the case of uncertainty visualisation would greatly help the field. (*XXX Is data pipeline connected with the grammar of graphics? Should this be a recommendation?*)

@@

If an uncertainty visualisation researcher would prefer to perform experiments rather than formalise methods, there are options there too. It would be interesting to know if any perceptual tasks that can be mapped to two different visual tasks condense into a single dimension when looking for overarching signal in a plot. Alternatively, the task dependency many authors in uncertainty visualisation mention would be a useful direction to consider. It is clear that the the number of potential tasks that can be performed on a visualisation increases with with the number of observations. A single observation is limited to value extraction, two observations can be compared, multiple observations allow for shapes or global statistics to be extracted. The interaction between sample size and task is of particular interest to the uncertainty visualisation community, as uncertainty can be expressed through multiple observations using a sample, or through a single value using an error. Of course, this is limited by the fact that there also isn't a definition for what is a "task" and given the mess created by the lack of formalisation in uncertainty visualisation, it may be wise to formalise that concept before performing these experiments. @Amar2005 suggested a taxonomy for information visualisation based on the types of tasks we use visualisations for and suggest 10 "analytical primitives" that we can then map to visualisations, which could be a good starting point. Regardless, these are directions of research would be fruitful to the uncertainty visualisation community even if it appears on the surface to be research that is only beneficial to the "normal" visualisation community. (*XXX Not sure what this paragraph is recommending?*)

### 2024-04-01 writingdumpster.qmd
removed old paragraphs after Di edit

# Removed Sections

>>>>

# Removed Sections
## Defining Uncertainty
The absence of an encompassing definition of uncertainty is mentioned by every uncertainty visualisation directly [@Spiegelhalter2017; @Griethe2006] or indirectly by describing a myriad of ways it can be considered in the literature [@Kinkeldey2014; @Hullman2016], although it is never commented on as a source of the noise in the field.

In order to visualise uncertainty, it needs to be quantifiable [@Griethe2006; Leland2005], in order to quantify uncertainty, it needs to be mathematically defined. The uncertainty visualisation literature is completely awash with papers that define uncertainty using a vague encompassing phrase, (e.g. @utypo defined it as any deviation from complete determinism) but go on to quantify uncertainty as a PDF, error, or some other easily quantified mathematical object when it comes time to do the visualisation. This definition swap happens so subtly that it seems to go unnoticed by authors
